---
title: Prefix operators in R
author: ''
date: '2019-09-11'
slug: prefix-operators-in-r
categories:
  - code
tags:
  - syntax
editor_options: 
  chunk_output_type: console
banner: "img/blog-images/header-images/mutate-prefixed-infix.png"
description: "Turns out there's a neat trick to use R's infix operators with prefix notation! Let's demonstrate how..."
---

```{r setup, include=FALSE}
library("emo")
library("askpass")
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

<span style="color:purple">Fair warning: This post is about an interesting feature of the R language, it likely won't help in your data analysis, wrangling or dataviz... but that doesn't stop it from being at least quite interesting.</span>

<hr>

Earlier today we were spelunking through the `add_tally()` function from `{dplyr}` and came across this wonderful line of code:

```{r, eval=FALSE,echo=TRUE}
mutate(x, `:=`(!!name, !!n))
```

That's an example of using **prefix notation** for something that we would normally write with **infix notation**. Let's compare the two:

<div class="row">
<div class="col-xs-6">
**infix notation**
```{r, eval=FALSE,echo=TRUE}
mutate(x, !!name := !!n)
```
</div>

<div class="col-xs-6">
**prefix notation**
```{r, eval=FALSE,echo=TRUE}
mutate(x, `:=`(!!name, !!n))
```
</div>
</div>

Hadley Wickham makes clear in his [Advanced R book](http://adv-r.had.co.nz/Functions.html#infix-functions) that most functions in R are written in prefix notation, e.g. `paste("hello", "world")` and `rep("hello world", 4)`. There are only a small number of built in functions that have a infix operator, and most of them are [arithmetic operators](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Arithmetic.html). 

Now we know how to re-write an infix operator in prefix notation, we can do something fairly special; we can swap the operator programmatically. Let's create a function where we can change the arithmetic operator applied to the first two [weird numbers](https://en.wikipedia.org/wiki/Weird_number):

```{r, echo=TRUE}
swappable_operator <- function(prefix){prefix(70, 836)}
```

Now we can give any of the infix operators as the argument for our function:

```{r, echo=TRUE, eval=FALSE}
swappable_operator(`+`)
## [1] 906
swappable_operator(`-`)
## [1] -766
swappable_operator(`*`)
## [1] 58520
```

That's quite fun, but there's not that much benefit to writing R code like this. However, in some other programming languages the ability to use prefix notation is incredibly useful. From 2012 - 2015 I was a Mathematica consultant for Wolfram Research and would use the following construction at least every other day:

```{wolframscript, echo=TRUE}
(*This is Wolfram Language code, NOT R*)
Accumulate@{1,2,4,16}
```

If anyone does have a good application for this prefix operator, please do let me know by replying to my tweet:

<center>
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Did you know that <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> has prefix notation via the backtick?<br><br>We only just discovered this today when spelunking through the code for add_tally() in {dplyr} 🧐 <a href="https://t.co/doRXUnlu24">pic.twitter.com/doRXUnlu24</a></p>&mdash; finding youR way (@rfindingyourway) <a href="https://twitter.com/rfindingyourway/status/1171758309398986757?ref_src=twsrc%5Etfw">September 11, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
</center>