---
title: 'juxtaposer: use cases and feature requirements'
author: ''
date: '2019-10-11'
slug: juxtaposer-use-cases
draft: yes
categories:
  - package development
  - htmlwidgets
  - dataviz
tags:
  - package development
  - juxtaposer
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library("tidyverse")
library("kableExtra")
```

I've been thinking about building the `{juxtapose}` package for awhile, and have put together enough **use cases** and **feature requirements** that I'm confident the package should be a thing. In this post I'm going to breakdown these use cases and features in two ways:

- The usual "wouldn't this be useful for an R user" way
- The [business analyst way](#enter-the-business-analysts)... read this last, or not at all.

### Use cases

#### Comparing photographs

  - Often journalists use JanitorJS to compare time periods by allowing readers to slide between two different photographs, here's a great example comparing [war torn Berlin in 1945 and modern day (2015)](https://interaktiv.morgenpost.de/berlin-1945-2015/).
  
  - Did you know that originally Ancient Roman statues were painted brightly by their creators? I highly recommend you read the excellent New Yorker article on ["the myth of whiteness in classical scripcture"](https://www.newyorker.com/magazine/2018/10/29/the-myth-of-whiteness-in-classical-sculpture)... sadly they didn't use JuxtaposeJS in their article, but look how impressive it could've been:  

<!--html_preserve-->
<div>
<link rel="stylesheet" href="https://cdn.knightlab.com/libs/juxtapose/latest/css/juxtapose.css">
<script src="https://cdn.knightlab.com/libs/juxtapose/latest/js/juxtapose.min.js"></script>
<link rel="stylesheet" href="https://cdn.knightlab.com/libs/juxtapose/latest/css/juxtapose.css">
<center>
<div id="recoloured-statues" style="width:200px">
</div>
</center>

<script>
    slider = new juxtapose.JXSlider('#recoloured-statues',
        [{
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/ancient-statue_without-colours.jpg",
                label: 'Present day'
            },
            {
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/ancient-statue_reconstructed-colours.jpg",
                label: 'Reconstructed colours'
            }
        ], {
            animate: true,
            showLabels: true,
            showCredits: false,
            startingPosition: "50%",
            makeResponsive: true
        });

</script>
</div>

<!--/html_preserve-->

#### Comparing dataviz

<!--html_preserve-->

<div class="row">

<div class="col-md-8">

<div class="well">

Choosing a good colour scale is difficult. Not only must the scale work well conceptually, it also needs to be readable by those with colour blindness.
<br>
This vertically-orientated JuxtaposeJS widget compares <a href="https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html" target="_blank">the most widely used colour scales</a> for a viewer with trichromacy [no colour blindness] and another viewer with Green-blind deuteranopia.

</div>

</div>

<div class="col-md-4">

<center>
<div id="colour-blindness" style="width:100%;max-width:300px">
</div>
</center>

</div>

</div>

<script>
    slider = new juxtapose.JXSlider('#colour-blindness',
        [{
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/colour-scales_trichromacy.png",
                label: 'Trichromacy'
            },
            {
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/colour-scales_deuteranopia.png",
                label: 'Deuteranopia'
            }
        ], {
            animate: true,
            showLabels: true,
            showCredits: false,
            mode: "vertical",
            startingPosition: "50%",
            makeResponsive: true
        });

</script>
</div>

<br>

<div class="row">

<div class="col-md-6">

<div class="well">

To claim that there are many ways to build a chart in R is an understatement. 
<br>
It's often useful to directly compare two different techniques side-by-side... and what better way than with JuxtaposeJS?

</div>

</div>

<div class="col-md-6">

<center>
<div id="charting-packages" style="width:100%;max-width:400px">
</div>
</center>

</div>

</div>

<script>
    slider = new juxtapose.JXSlider('#charting-packages',
        [{
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/scatter-plot_base-r.png",
                label: 'Base R plot()'
            },
            {
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/scatter-plot_ggplot2.png",
                label: '{ggplot2} chart'
            }
        ], {
            animate: true,
            showLabels: true,
            showCredits: false,
            startingPosition: "50%",
            makeResponsive: true
        });

</script>

<!--/html_preserve-->

#### Comparing code (in training slides)

<!--html_preserve-->

<div class="row">

<div class="col-md-6">

<div class="well">

Often when writing code tutorials it's convenient to compare one approach to another.

I'm a great fan of Hadley Wickham's <a href="https://twitter.com/AmeliaMN/status/748193609401327616" target="_blank">little bunny foo foo example</a> for demonstrating the advantages to using pipes.
<br>
<br>
This <i>would</i> work in a JuxtaposeJS element, but only if we changed the colour of the slider.
</div>

</div>

<div class="col-md-6">

<center>
<div id="code-comparison" style="width:100%;max-width:400px">
</div>
</center>

</div>

</div>

<script>
    slider = new juxtapose.JXSlider('#code-comparison',
        [{
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/foo-foo_traditional-notation.png",
            },
            {
                src: "/blog/2019-10-11-juxtaposer-use-cases_files/foo-foo_pipes.png",
            }
        ], {
            animate: true,
            showCredits: false,
            startingPosition: "50%",
            makeResponsive: true
        });

</script>

<!--/html_preserve-->

### ... enter the Business Analysts {#enter-the-business-analysts}

Several years ago I worked as a buisness analyst (BA), and really enjoyed the job. BAs are the folks who are responsible for **requirements gathering**, **maturing requirement**, and bridging the gap between **client expectations vs developer resources for new features**.

It seems to me that the best R packages are built with a BA mindset:

- Great packages are built with a specific idea in mind, example: `{ggplot2}` was designed to provide a consistent grammar of graphics to R users.

- Amazing devs take feedback and usage data to mature the package, example: `{ggplot2}`'s internals have been optimised continuously to improve rendering speed.

- Awesome Devs decide how much they can extend their packages, example: `{ggplot2}` now has support for `sf` objects and a matured API for extensions, for instance `{ggrepel}`

So, let's consider how we might approach this package as a BA... to do that, we need to pretend to have the following colleagues, all vying for the attention of our dev time:

- the client/customer (the R user community)
- the project manager (the time commitments of your other projects)
- the developer (the time you can commit to this project)

### Requirements gathering

We need to have a common language for talking to all of our colleagues, particularly when it comeas to deciding which are the features which **must** be in the CRAN release of our package.

In BA speak, the core requirements are contained within **user stories**, for example:

1. A user views a `juxtaposer` htmlwidget in an RMarkdown document

1. A user views a `juxtaposer` htmlwidget in a Shiny app

1. A user uploads their own image(s) to a Shiny app which is then viewed in a `juxtapose` htmlwidget.

Before I write out all of the requirements, let me introduce a useful BA tool for feature prioritisation: [MoSCow](https://en.wikipedia.org/wiki/MoSCoW_method). For this package, we're going to label each requirement as a **must**, **should**, **could** or **won't** by following this guide:

```{r, echo=FALSE}
moscow <- tribble(
  ~`moscow`, ~`Description`,
  "Must have", "The feature must be in the devel release of the package",
  "Should have", "The feeature must be in the first CRAN release of the package",
  "Could have", "Scaffolding for featue should be considered before the first CRAN release, to reduce the likelihood of significant API changes in future releases",
  "Won't have", "The feature will be added to the issues log as an \"enhancement\""
)

moscow %>%
  kable(col.names = c("MoSCoW", "Description")) %>%
  kable_styling()

```

Now, finally here are the requirements as I see them:

```{r, echo=FALSE}
juxtaposer_requirements <- tribble(
  ~user_story, ~moscow, ~dev_notes,
  "User views a `{juxtaposer}` widget in an RMarkdown document", "Must", "This comes for free thanks to the `{htmlwidget}` package, it'll just work.",
  "User views a `{juxtaposer}` widget in a Shiny app", "Must", "This comes for free thanks to the `{htmlwidget}` package, it'll just work.",
  "Dev can set custom colours for the slider elements in a `{juxtaposer}` widget", "Must", "<p>The best approach for this would be to follow the design of `kableExtra::kable_Styling()`.</p><p>This is because styles are set in CSS instead of in JavaScript, see <a href='https://github.com/charliejhadley/juxtaposer/issues/2' target='_blank'>Issue #2</a> for more details.</p>",
  "A user clicks a control in a Shiny app which results in the images in the `{juxtaposer}` widget changing.", "Should", "<p>Most htmlwidgets allow widgets to be updated via a proxy, which also feels like the best option here.</p><p>As well as updating the images, the dev should also be able to change the slider colours in case they contrast with the new images. See <a href='https://github.com/charliejhadley/juxtaposer/issues/3' target='_blank'>Issue #3</a> for details.</p>",
  "A user clicks a control in a Shiny app which results in the position of the `{juxtaposer}` slider changing", "Could",
  "Provided `juxtaposer_proxy()` has already been implememnted, we need to extend the package with `juxtaposer_update_slider()`. See <a href='https://github.com/charliejhadley/juxtaposer/issues/4' target='_blank'>Issue 4</a> for details.",
  "User uploads two new images to a `{juxtaposer}` widget via the `juxtaposerInput() control`", "Won't", "There's an initial sketch of the functionality in <a href='https://github.com/charliejhadley/juxtaposer/issues/5' target=''>Issue 5</a>."
)

juxtaposer_requirements %>%
  kable(col.names = c("User story", "MoSCoW", "Dev Notes"),
        escape = FALSE) %>%
  kable_styling()
```

